---
title: "NDD Racial Disparity Dashboard"
author: "Jiayi Liu"
output: 
  flexdashboard::flex_dashboard:
    orientation: columns
    vertical_layout: fill
    source_code: "https://github.com/Jiayi-Liu7/NDD_Racial_Disparity"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(tidyverse)
library(janitor)
library(stringr)
library(scales)
library(dplyr)
library(ggplot2)
library(plotly)
```

# Dashboard {data-icon="fa-signal"}

```{r, include=FALSE}
# Note: The raw EHC dataset is confidential. Please ensure the dataset is in your working directory.
datapath <- "Racial Disparity Phase 2 - 03.04.2026.csv"

df <- read_csv(datapath, show_col_types = FALSE) %>%
  clean_names()

df <- df %>%
  mutate(
    exclude = str_squish(as.character(exclude)),
    exclude = na_if(exclude, "")   # treat blanks as NA
  )

# Apply manual exclusions (Exclude == "x")
df <- df %>%
  filter(is.na(exclude) | tolower(exclude) != "x")

# Create person-level dataset with Case vs Control
df_case <- df %>%
  transmute(
    pair_id, row_id, case_fac = "Case",
    dx_cd, dx_desc, race = race_desc, gender = gender_desc, age = age_at_visit,
    dx_provider_nm, dx_provider_dept, prov_name_closest_visit, prov_spec_closest_visit,
    mons_btw_dx_pcp, emory_assoc_pcp, closest_visit_prov_pcp_flag
  )

df_control <- df %>%
  transmute(
    pair_id, row_id = row_id_cont, case_fac = "Control",
    dx_cd = NA_character_, dx_desc = NA_character_, race = cont_race_desc,
    gender = cont_gender_desc, age = cont_age_at_visit,
    cont_provider_nm, cont_provider_spec,
    dx_provider_nm, dx_provider_dept, prov_name_closest_visit, prov_spec_closest_visit,
    mons_btw_dx_pcp, emory_assoc_pcp, closest_visit_prov_pcp_flag
  )

df_long <- bind_rows(df_case, df_control) %>%
  mutate(
    case_fac = factor(case_fac, levels = c("Control", "Case")),
    case_indicator = if_else(case_fac == "Case", 1L, 0L)
  )

# Variables recoding
df_long <- df_long %>%
  mutate(
    race = str_squish(as.character(race)),
    race_grouped = case_when(
      race %in% c("Chinese", "Other Asian", "Asian Indian", "Korean", "Vietnamese", "Other Pacific Islander", "Native Hawaiian") ~ "Asian",
      race %in% c("American Indian or Alaska Native", "American Indian or Alaskan Native", "Multiple", "Other") ~ "Other/Small Group",
      race %in% c("Decline to Answer", "Unknown", "Not Recorded") | is.na(race) | race == "" ~ "Unknown/Not Reported",
      race %in% c("White", "Caucasian or White") ~ "White",
      TRUE ~ race
    ),
    race_grouped = factor(race_grouped, levels = c("White", "Black or African American", "Asian", "Other/Small Group", "Unknown/Not Reported"))
  )

# Disease group mapping
df_long <- df_long %>%
  mutate(
    dx_desc_clean = if_else(is.na(dx_desc), NA_character_, str_squish(as.character(dx_desc))),
    dx_cd_raw = if_else(is.na(dx_cd), NA_character_, str_to_upper(str_squish(as.character(dx_cd)))),
    dx_cd_raw = if_else(is.na(dx_cd_raw), NA_character_, str_replace_all(dx_cd_raw, "\\s+", "")),
    dx_cd_raw = if_else(is.na(dx_cd_raw), NA_character_, str_replace_all(dx_cd_raw, "[^A-Z0-9.]", "")),
    
    dx_cd_std = case_when(
      is.na(dx_cd_raw) ~ NA_character_,
      str_starts(dx_cd_raw, "G") ~ dx_cd_raw,
      TRUE ~ str_replace(dx_cd_raw, "(\\.[0-9]*?)0+$", "\\1") %>% str_replace("\\.$", "")
    ),
    
    disease_group = case_when(
      case_fac != "Case" ~ NA_character_,
      dx_cd_std == "G30.0" ~ "EOAD",
      dx_cd_std == "G30.1" ~ "LOAD",
      dx_cd_std %in% c("G30.9", "331", "G30.8") & age >= 65 ~ "LOAD",
      dx_cd_std %in% c("G30.9", "331", "G30.8") & age < 65 ~ "EOAD",
      dx_cd_std %in% c("G20", "332") ~ "PD",
      dx_cd_std %in% c("G31.83", "331.82") ~ "DLB",
      dx_cd_std %in% c("335.2", "G12.21", "G12.22", "G12.20", "G12.29", "335.24") ~ "MND",
      dx_cd_std %in% c("G31.09", "331.19", "G31.01") ~ "FTD",
      TRUE ~ NA_character_
    ),
    disease_group = factor(disease_group, levels = c("LOAD", "EOAD", "FTD", "DLB", "PD", "MND"))
  )

# Match Control to Case Disease
case_disease <- df_long %>%
  filter(case_fac == "Case", !is.na(disease_group)) %>%
  distinct(pair_id, disease_group)

df_disease_cc <- df_long %>%
  left_join(case_disease, by = "pair_id", suffix = c("", "_case"), relationship = "many-to-many") %>%
  mutate(
    disease_group_case = factor(disease_group_case, levels = levels(df_long$disease_group)),
    case_fac = factor(case_fac, levels = c("Control", "Case")),
    race_grouped = fct_drop(race_grouped)
  ) %>%
  filter(!is.na(disease_group_case))
```


Column {data-width=400}
-----------------------------------------------------------------------

### Widget 1: Racial Composition Across Neurodegenerative Diseases

```{r}
# Prepare data for Widget 1
vis1_data <- df_long %>%
  count(case_fac, race_grouped) %>%
  group_by(case_fac) %>%
  mutate(
    prop = n / sum(n),
    prop_rounded = round(prop * 100, 1),
    tooltip_text = paste0(
      "Group: ", case_fac, "\n",
      "Race: ", race_grouped, "\n",
      "Count: ", n, " participants\n",
      "Proportion: ", prop_rounded, "%"
    )
  ) %>%
  ungroup()

# Create ggplot
p1 <- ggplot(vis1_data, aes(x = case_fac, y = prop, fill = race_grouped, text = tooltip_text)) +
  geom_col(position = "fill") +
  theme_bw(base_size = 13) +
  labs(
    title = "Race Distribution of Cases vs Controls",
    x = "Patient Group (Control vs. Case)",
    y = "Proportion of Participants (%)",
    fill = "Race Group"
  ) +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_brewer(palette = "Set2")

# Convert to interactive plotly widget
ggplotly(p1, tooltip = "text") %>%
  layout(hoverlabel = list(bgcolor = "white"))
```


Column {data-width=600}
-----------------------------------------------------------------------

### Widget 2: White vs Black Patient Composition by Disease Group

```{r}
# Prepare data for Widget 2
plot_slope_white_black <- df_disease_cc %>%
  filter(
    !is.na(disease_group_case),
    race_grouped %in% c("White", "Black or African American")
  ) %>%
  count(disease_group_case, case_fac, race_grouped, name = "n") %>%
  group_by(disease_group_case, case_fac) %>%
  mutate(
    pct = n / sum(n),
    pct_rounded = round(pct * 100, 1),
    tooltip_text = paste0(
      "Disease: ", disease_group_case, "\n",
      "Race: ", race_grouped, "\n",
      "Group: ", case_fac, "\n",
      "Count: ", n, "\n",
      "Proportion: ", pct_rounded, "%"
    )
  ) %>%
  ungroup() %>%
  mutate(
    case_fac = factor(case_fac, levels = c("Control", "Case")),
    race_grouped = factor(race_grouped, levels = c("White", "Black or African American"))
  )

# Create ggplot
p2 <- ggplot(plot_slope_white_black, aes(x = case_fac, y = pct, group = race_grouped, color = race_grouped, text = tooltip_text)) +
  geom_line(linewidth = 1) +
  geom_point(size = 3) +
  facet_wrap(~ disease_group_case, ncol = 3) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  theme_bw(base_size = 12) +
  theme(
    axis.title.x = element_blank(),
    legend.position = "bottom"
  ) +
  labs(
    title = "White vs Black Patient Composition by Disease Group",
    y = "Percent within Disease Group (%)",
    color = "Race Group"
  )

# Convert to interactive plotly widget
ggplotly(p2, tooltip = "text") %>%
  layout(legend = list(orientation = "h", x = 0.2, y = -0.15))
```

### Real-world Impact & Importance

This dashboard visualizes racial disparities in neurodegenerative disease diagnoses using matched case-control data to highlight systemic inequities. By identifying specific conditions where disparities persist, these insights can directly guide targeted clinical interventions and health policy reforms.


# Dataset Description {data-icon="fa-info-circle"}

Column {data-width=1000}
-----------------------------------------------------------------------

### Data Source and Methodology

This project utilizes a confidential clinical public health dataset from the Emory Healthcare system, containing approximately 10,500 patient records spanning from 2015 to 2024. The analysis employs a 1-to-1 matched case-control design to compare demographic representation, specifically racial composition, across various neurodegenerative conditions. Due to patient privacy regulations and HIPAA compliance, the raw individual-level dataset is restricted and not included in the public repository. All visualizations presented in this dashboard are generated from aggregated, de-identified metrics. This ensures data security while accurately reflecting validated clinical registry outcomes.