---
title: "Preliminary result"
format: html
editor: visual
---

## Load data

```{r, echo=TRUE, output=FALSE}
library(tidyverse)
library(janitor)
library(stringr)
library(scales)
library(gtsummary)
library(dplyr)
library(ggplot2)

# Note: The raw EHC dataset is confidential. Please ensure the dataset is in your working directory.
datapath <- "Racial Disparity Phase 2 - 03.04.2026.csv"

df <- read_csv(datapath, show_col_types = FALSE) %>%
  clean_names()

df <- df %>%
  mutate(
    exclude = str_squish(as.character(exclude)),
    exclude = na_if(exclude, "")   # treat blanks as NA
  )

# Apply manual exclusions (Exclude == "x")
df <- df %>%
  filter(is.na(exclude) | tolower(exclude) != "x")
```

### Create person-level dataset with Case vs Control

Because each row is a case–control pair, we now split into two datasets.

Now it's one row per person with variables `age`, `race`, `gender`, and a factor `case_fac` indicating whether the row is a case or a control.

```{r}
df_case <- df %>%
  transmute(
    pair_id,
    row_id,
    case_fac = "Case",
    dx_cd,
    dx_desc,
    race   = race_desc,
    gender = gender_desc,
    age    = age_at_visit,
    
    # --- provider-related fields (CASE) ---
    dx_provider_nm,
    dx_provider_dept,
    prov_name_closest_visit,
    prov_spec_closest_visit,
    mons_btw_dx_pcp,
    emory_assoc_pcp,
    closest_visit_prov_pcp_flag
  )

df_control <- df %>%
  transmute(
    pair_id,
    row_id = row_id_cont,
    case_fac = "Control",
    dx_cd = NA_character_,
    dx_desc = NA_character_,
    race   = cont_race_desc,
    gender = cont_gender_desc,
    age    = cont_age_at_visit,
    
    # --- provider-related fields (CONTROL) ---
    cont_provider_nm,
    cont_provider_spec,

    # case-derived fields
    dx_provider_nm,
    dx_provider_dept,
    prov_name_closest_visit,
    prov_spec_closest_visit,
    mons_btw_dx_pcp,
    emory_assoc_pcp,
    closest_visit_prov_pcp_flag
  )

df_long <- bind_rows(df_case, df_control) %>%
  mutate(
    case_fac = factor(case_fac, levels = c("Control", "Case")),
    case_indicator = if_else(case_fac == "Case", 1L, 0L)
  )
```

### 

## Race recoding and validation

Several race categories have very small sample sizes (e.g., American Indian or Alaska Native, Chinese, Other). So we collapse them into 5 broader groups:

-   White/Caucasian or White

-   Black or African American

-   Asian

-   Other/Small Group

-   Unknown/Not Reported - Decline to Answer, Unknown, Not Recorded

```{r}
df_long <- df_long %>%
  mutate(
    race = str_squish(as.character(race)),
    race_grouped = case_when(
      # Asian categories
      race %in% c("Chinese", "Other Asian", "Asian Indian", "Korean", "Vietnamese",
                  "Other Pacific Islander", "Native Hawaiian") ~ "Asian",

      # Small minority categories
      race %in% c("American Indian or Alaska Native",
                  "American Indian or Alaskan Native",
                  "Multiple",
                  "Other") ~ "Other/Small Group",

      # Missing / unclear
      race %in% c("Decline to Answer", "Unknown", "Not Recorded") ~ "Unknown/Not Reported",
      is.na(race) | race == "" ~ "Unknown/Not Reported",

      # White
      race %in% c("White", "Caucasian or White") ~ "White",
      
      # Black
      TRUE ~ race
    ),
    race_grouped = factor(
      race_grouped,
      levels = c("White", "Black or African American",
                 "Asian", "Other/Small Group", "Unknown/Not Reported")
    )
  )
```

1.  Show sample size for each grouped race category

```{r}
df_long %>%
  count(race_grouped) %>%
  mutate(percent = round(100 * n / sum(n), 1)) %>%
  arrange(desc(n))
```

2.  Show case vs control counts within each race group

```{r}
df_long %>% count(race_grouped, case_fac) %>% arrange(race_grouped)
```

## Disease-group mapping

```{r}
df_long <- df_long %>%
  mutate(
    dx_desc_clean = if_else(is.na(dx_desc), NA_character_, str_squish(as.character(dx_desc))),

    # clean DX_CD
    dx_cd_raw = if_else(is.na(dx_cd), NA_character_, str_to_upper(str_squish(as.character(dx_cd)))),
    dx_cd_raw = if_else(is.na(dx_cd_raw), NA_character_, str_replace_all(dx_cd_raw, "\\s+", "")),
    dx_cd_raw = if_else(is.na(dx_cd_raw), NA_character_, str_replace_all(dx_cd_raw, "[^A-Z0-9\\.]", "")),

    # If starts with 'G' (ICD-10 style), keep dot + trailing 0 (e.g., G30.0)
    # Otherwise (mostly numeric ICD-9 style), trim trailing zeros after decimal (e.g., 335.20 -> 335.2)
    dx_cd_std = case_when(
      is.na(dx_cd_raw) ~ NA_character_,
      str_starts(dx_cd_raw, "G") ~ dx_cd_raw,
      TRUE ~ str_replace(dx_cd_raw, "(\\.[0-9]*?)0+$", "\\1") |> str_replace("\\.$", "")
    ),

    disease_group = case_when(
      case_fac != "Case" ~ NA_character_,

      # --- AD ---
      dx_cd_std == "G30.0" ~ "EOAD",
      dx_cd_std == "G30.1" ~ "LOAD",
      dx_cd_std %in% c("G30.9", "331", "G30.8") & age >= 65 ~ "LOAD",
      dx_cd_std %in% c("G30.9", "331", "G30.8") & age <  65 ~ "EOAD",

      # --- PD ---
      dx_cd_std %in% c("G20", "332") ~ "PD",

      # --- DLB ---
      dx_cd_std %in% c("G31.83", "331.82") ~ "DLB",

      # --- MND ---
      dx_cd_std %in% c("335.2", "G12.21", "G12.22", "G12.20", "G12.29", "335.24") ~ "MND",

      # --- FTD ---
      dx_cd_std %in% c("G31.09", "331.19", "G31.01") ~ "FTD",

      TRUE ~ NA_character_
    ),

    disease_group = factor(disease_group, levels = c("LOAD", "EOAD", "FTD", "DLB", "PD", "MND"))
  )
```

Diagnoses were grouped into six disease categories based on observed diagnosis descriptions in the Emory dataset. Alzheimer’s disease was further stratified into late-onset (LOAD) and early-onset (EOAD) using age at diagnosis. Diagnoses that did not map to these categories were excluded from disease-stratified analyses.

### Diagnosis count table

```{r}
df_long %>%
  filter(case_fac == "Case") %>%
  count(disease_group, dx_desc_clean, sort = TRUE)
```

## Table 1

Table 1 summarizes baseline demographic characteristics for cases and controls at the person level.

```{r}
table1 <- df_long %>%
  select(case_fac, age, gender, race_grouped) %>%
  tbl_summary(
    by = case_fac,
    statistic = list(
      all_continuous() ~ "{mean} ({sd})",
      all_categorical() ~ "{n} ({p}%)"
    ),
    missing = "no",
    label = list(
      age ~ "Age at Visit",
      gender ~ "Gender",
      race_grouped ~ "Race Group"
    )
  ) %>%
  add_overall() %>%
  add_p() %>%
  modify_caption("**Table 1. Baseline Demographic Characteristics of Cases and Controls**")

table1
```

## Age distribution across races for all participants

```{r}
ggplot(df_long, aes(x = race_grouped, y = age, fill = race_grouped)) +
  geom_boxplot(alpha = 0.7) +
  theme_bw(base_size = 12) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1)) +
  labs(title = "Age Distribution Across Race Group",
       x = "Race",
       y = "Age at Visit") +
  scale_fill_brewer(palette = "Set2")
```

## Age density by race for all participants

After collapsing very small racial categories into broader groups (Other/Small Group), the age density curves became more stable and less spiky.

```{r}
ggplot(df_long, aes(x = age, color = race_grouped, fill = race_grouped)) +
  geom_density(alpha = 0.20, adjust = 2) +
  theme_bw(base_size = 13) +
  labs(title = "Age Density by Race Group",
       x = "Age at Visit",
       y = "Density")
```

## Gender distribution by race for all participants

For each race group, the bar shows the proportion of males and females (bars sum to 100% within race).

```{r}
df_long %>%
  count(race_grouped, gender) %>%
  group_by(race_grouped) %>%
  mutate(prop = n/sum(n)) %>%
  ggplot(aes(x = race_grouped, y = prop, fill = gender)) +
  geom_col() +
  geom_text(aes(label = scales::percent(prop, accuracy = 1)),
            position = position_stack(vjust = 0.5),
            size = 3) +
  theme_bw(base_size = 13) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1)) +
  labs(title = "Gender Distribution Within Each Race Group",
       x = "Race",
       y = "Proportion") +
  scale_y_continuous(labels = scales::percent)
```

```{r}
# Gender distribution by race (counts)
df_long %>%
  count(race_grouped, gender) %>%
  ggplot(aes(x = race_grouped, y = n, fill = gender)) +
  geom_col() +
  geom_text(
    aes(label = n),
    position = position_stack(vjust = 0.5),
    size = 3) +
  theme_bw(base_size = 13) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1)) +
  labs(
    title = "Gender Distribution by Race Group (Counts)",
    x = "Race",
    y = "Number of Participants",
    fill = "Gender"
  )
```

## Age distribution in Cases vs Controls

```{r}
ggplot(df_long, aes(x = case_fac, y = age, fill = case_fac)) +
  geom_boxplot(alpha = 0.7) +
  theme_bw(base_size = 13) +
  labs(
    title = "Age Distribution of Cases vs Control",
    x = "Group",
    y = "Age at Visit"
  ) +
  guides(fill = "none")
```

```{r}
ggplot(df_long, aes(x = case_fac, y = age)) +
  geom_boxplot(alpha = 0.7) +
  facet_wrap(~ race_grouped) +
  theme_bw(base_size = 12) +
  labs(
    title = "Age Distribution of Cases vs Controls Within Race Groups",
    x = "Group",
    y = "Age at Visit"
  )

```

## Race distribution in Cases vs Controls

```{r}
df_long %>%
  count(case_fac, race_grouped) %>%
  group_by(case_fac) %>%
  mutate(prop = n/sum(n)) %>%
  ggplot(aes(x = case_fac, y = prop, fill = race_grouped)) +
  geom_col() + 
  geom_text(aes(label = scales::percent(prop, accuracy = 1)),
            position = position_stack(vjust = 0.5),
            size = 3) + 
  theme_bw(base_size = 13) +
  labs(title = "Race Distribution of Cases vs Controls",
       x = "Group",
       y = "Proportion") +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_brewer(palette = "Set2")
```

## Disease matched case/control dataset

```{r}
# Assign controls the disease group of their matched case
case_disease <- df_long %>%
  filter(case_fac == "Case", !is.na(disease_group)) %>%
  distinct(pair_id, disease_group)

df_disease_cc <- df_long %>%
  left_join(case_disease, by = "pair_id", suffix = c("", "_case")) %>%
  mutate(
    disease_group_case = disease_group_case,
    disease_group_case = factor(disease_group_case, levels = levels(df_long$disease_group)),
    case_fac = factor(case_fac, levels = c("Control", "Case")),
    race_grouped = fct_drop(race_grouped)   # drop unused levels (helps plotting)
  ) %>%
  filter(!is.na(disease_group_case))
```

### Table: Case vs Control counts by disease group

```{r}
df_disease_cc %>%
  count(disease_group_case, case_fac) %>%
  group_by(disease_group_case) %>%
  mutate(percent = round(100 * n / sum(n), 1)) %>%
  ungroup()
```

### Race composition by disease group

```{r}
# Race composition by disease group
plot_slope_white_black <- df_disease_cc %>%
  filter(
    !is.na(disease_group_case),
    race_grouped %in% c("White", "Black or African American")
  ) %>%
  count(disease_group_case, case_fac, race_grouped, name = "n") %>%
  group_by(disease_group_case, case_fac) %>%
  mutate(pct = n / sum(n)) %>%
  ungroup() %>%
  mutate(
    case_fac = factor(case_fac, levels = c("Control", "Case")),
    race_grouped = factor(
      race_grouped,
      levels = c("White", "Black or African American")
    )
  )

ggplot(plot_slope_white_black,
       aes(x = case_fac, y = pct, group = race_grouped, color = race_grouped)) +
  geom_line(linewidth = 1.4) +
  geom_point(size = 3) +

  geom_text(
    data = subset(plot_slope_white_black, case_fac == "Control"),
    aes(label = percent(pct, accuracy = 0.1)),
    hjust = 1.1, vjust = 0.5, size = 3.2, show.legend = FALSE
  ) +
  geom_text(
    data = subset(plot_slope_white_black, case_fac == "Case"),
    aes(label = percent(pct, accuracy = 0.1)),
    hjust = -0.1, vjust = 0.5, size = 3.2, show.legend = FALSE
  ) +

  facet_wrap(~ disease_group_case, ncol = 3) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  coord_cartesian(clip = "off") +
  theme_bw(base_size = 12) +
  theme(
    plot.margin = margin(8, 60, 8, 60),
    axis.title.x = element_blank(),
    legend.position = "bottom"
  ) +
  labs(
    title = "White vs Black/African American composition by disease group",
    y = "Percent within disease group",
    color = "Race group"
  )
```

```{r}
# Table 2. Baseline demographics by disease group (cases) + Control
df_table2 <- df_long %>%
  mutate(
    disease_group_cc = if_else(case_fac == "Control", "Control", as.character(disease_group))
  ) %>%
  filter(!is.na(disease_group_cc)) %>%
  mutate(
    disease_group_cc = factor(
      disease_group_cc,
      levels = c("Control", levels(df_long$disease_group))
    )
  )

table2 <- df_table2 %>%
  select(disease_group_cc, age, gender, race_grouped) %>%
  tbl_summary(
    by = disease_group_cc,
    statistic = list(
      all_continuous() ~ "{mean} ({sd})",
      all_categorical() ~ "{n} ({p}%)"
    ),
    missing = "no",
    label = list(
      age ~ "Age at Visit",
      gender ~ "Gender",
      race_grouped ~ "Race Group"
    )
  ) %>%
  add_overall() %>%
  add_p() %>%
  modify_caption("**Table 2. Baseline Demographic Characteristics by Disease Group (Cases) and Controls**")

table2
```

EOAD Primary care provider

```{r}
df_eoad_cc <- df_disease_cc %>%
  filter(disease_group_case == "EOAD") %>%
  mutate(
    case_fac = str_squish(as.character(case_fac))
  )

eoad_pcp_provider_long <- df_eoad_cc %>%
  mutate(
    primary_care_provider = case_when(
      case_fac == "Case"    ~ prov_name_closest_visit,
      case_fac == "Control" ~ cont_provider_nm,
      TRUE ~ NA_character_
    ),
    primary_care_provider = str_squish(as.character(primary_care_provider))
  ) %>%
  filter(!is.na(primary_care_provider), primary_care_provider != "") %>%
  count(primary_care_provider, case_fac, name = "n") %>%
  arrange(desc(n))

eoad_pcp_provider_long
```

EOAD Primary care location

```{r}
eoad_pcp_specialty <- df_eoad_cc %>%
  mutate(
    primary_care_location = case_when(
      case_fac == "Case"    ~ prov_spec_closest_visit,
      case_fac == "Control" ~ cont_provider_spec,
      TRUE ~ NA_character_
    ),
    primary_care_location = str_squish(as.character(primary_care_location))
  ) %>%
  filter(!is.na(primary_care_location), primary_care_location != "") %>%
  count(primary_care_location, case_fac, name = "n") %>%
  group_by(case_fac) %>%
  mutate(pct_within_group = n / sum(n)) %>%
  ungroup() %>%
  arrange(desc(n))

eoad_pcp_specialty
```

EOAD Diagnosis departments

```{r}
eoad_dx_dept_caseonly <- df_eoad_cc %>%
  filter(case_fac == "Case") %>%
  mutate(
    dx_provider_dept = str_squish(as.character(dx_provider_dept))
  ) %>%
  filter(!is.na(dx_provider_dept), dx_provider_dept != "") %>%
  count(dx_provider_dept, name = "n") %>%
  arrange(desc(n))

eoad_dx_dept_caseonly
```
